Nitrogen's form determines its fate in the North Pacific Subtropical Gyre during PARAGON I

### 2025-05-23 manuscript/manuscript.Rmd
New introduction drafted

Marine carbon is fixed into biomass very quickly. In most of the surface ocean this flux is controlled by the availability of nitrogen, an element with a complex and incompletely characterized biogeochemical cycle [@Capone2008; @Moore2013; @Hutchins2022]. Organic nitrogen in particular is often treated as a black box despite substantial variability in its bioavailability and chemical nature. While extensive work has been done to characterize the forms and fluxes of nitrogen within the inorganic pool, the equivalent for the organic pool is woefully underdeveloped [@ref, prob NME2008 or Hansell2024, maybe @Hutchins2022 for visual?]. This is largely due to the difficulty of comprehensively measuring the many organic molecules that contain nitrogen in the marine environment [@Boysen2018; @Moran2022]. The bioavailability of the nitrogen atom is in the environment is largely determined by the form of the organic material it composes.

The mechanisms by which dissolved nitrogen becomes biomass and vice versa are major factors in our ability to predict ecosystem productivity and therefore carbon fixation and export. Traditionally, the conversion of inorganic material into organic substrate was thought to be limited to autotrophs while remineralization was performed by the heterotrophic community. Today, it’s known that many photosynthetic organisms are able to take up and use organic nitrogen [@Anita1991; @Morando2018; @Hugo2021] and that only a select subset of phototrophs are capable of nitrate reduction or nitrogen fixation [@ref], while inorganic nitrogen is directly accessible to heterotrophs as well. This means that the transformations possible for a given N-containing compound are a function of the microbial community and its environment while the microbial community is in turn also expected to be a function of the N-containing molecules available. This creates a recursive network that’s difficult to untangle and cannot necessarily be extrapolated from axenic culture studies.

One way to reveal the pathways and transformations that marine N undergoes in a natural community is via 15N labeling paired with metabolomics. Metabolites that incorporate the labeled nitrogen can then be separated on the mass spectrometer and quantified separately from the unlabeled pool. Here, we trace the uptake and use of labeled ammonia, nitrate, arginine, and guanosine monophosphate in a natural community from the NPSG at multiple depths and diel conditions. Our application of metabolomics to these samples allows us to map out the pathways and restructuring that organic nitrogen experiences in the largest biomes on the planet, quantifying the pool sizes and turnover rates for various low molecular weight compounds that serve as building blocks and intermediates of cell biology.

@@

>>>>

In most of the surface ocean the flux of carbon is controlled by the availability of nitrogen, an element with a complex and incompletely characterized biogeochemical cycle [@Capone2008; @Moore2013; @Hutchins2022]. Organic nitrogen in particular is often treated as a black box despite substantial variability in its bioavailability and chemical nature. Turnover times for labile organic compounds like urea and amino acids can be orders of magnitude shorter than high molecular weight DON [@Aluwihare2005; @Bronk2024]. Additionally, the availability of the precious nitrogen atom is determined not only by the form of the organic material it composes but also the environment in which it's found.

While extensive work has been done to characterize the forms and fluxes of nitrogen within the inorganic pool, the equivalent for the organic pool is woefully underdeveloped [@Zehr2011; @NME]. This is largely due to the difficulty of comprehensively measuring the many organic molecules that contain nitrogen in the marine environment [@Boysen2018; @Moran2022]. Recently, however, we have improved our understanding of community composition via genomics and transcriptomics and made progress in our ability to characterize organic matter via proteomics and metabolomics. These advancements, paired with the increased availability of pure and isotope-labeled substrates, mean that we can now apply metabolically resolved isotope tracing to resolve how organic nitrogen is transformed in the marketplace of metabolites that empower the surface ocean.

Nitrogen assimilation is a well-documented process in the ocean after decades of research and demonstrates how biochemistry derived from non-marine organisms can be applied to the ocean. Central to the process is ammonia, which can be directly taken up from the environment by all organisms and combined with glutamate. This forms glutamine, which is then combined with alpha-ketoglutarate to form two glutamate molecules that can then be used as a nitrogen source throughout the cell [@Walker2016]. Nitrate, on the other hand, must first be reduced to ammonia in an energetically expensive process that some organisms have stripped from their genomes [@Partensky2010; @Zehr2011].

Unlike the relatively well-resolved pathways for inorganic nitrogen, the use of organic nitrogen remains underexplored in the marine environment. Organic nitrogen is generally thought to be accessible to most organisms but the high levels of DON make it clear that this is highly substrate-specific [@Antia1991; @Bronk2007; @Aluwihare2008; @Zehr2011; @Bronk2024]. Additionally, many of the transformations possible are only active in specific communities or times of day [@Bender2012]. Linking the transformations that nitrogen undergoes to known molecular pathways documented in other organisms or in culture is a particularly promising way to understand what we observe on the global scale.

The lack of a comprehensive understanding of nitrogen's fate after it is assimilated makes molecular modeling difficult and undermines our ability to connect the community to the organic nitrogen it produces. Here, we trace the uptake and use of labeled ammonia, nitrate, arginine, and guanosine monophosphate in a natural community from the North Pacific Subtropical Gyre at multiple depths and diel conditions. Our application of metabolomics to these samples allows us to map the pathways and restructuring that organic nitrogen experiences in the largest biomes on the planet, quantifying the pool sizes and turnover rates for various low molecular weight compounds that serve as building blocks and intermediates of cell biology.

@@

Incubations are imperfect metrics of the actual environment because amendments with *anything* will heavily skew the community - over what time scale though? [@Eppley1977; @Shilova2017; @Rii2018]
